Multi-reference pangenome graphs

### 2022-01-05 manuscript/main.tex
formally

@@

\section{Variation graph induction}
\label{sec:model}

A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
A \textit{bidirected} sequence graph represents both strands of DNA.
On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.

A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
A node strand $s$ corresponds to one node orientation.
Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.

A variation graph represents pairwise alignments between its embedded paths.
We infer that two sequences match when they both ``step'' on the same node or series of nodes.
This follows from our definition of the graph itself.
If two sub-paths contain the same set of steps, they must have identical subsequences.
%Thus, to represent alignments in the graph, we built it by rendering matches into the graph
%The core of our graph constr TODO HERE
We can construct a variation graph from a set of pairwise alignments by guaranteeing that each matching pair of subsequences are represented in a single walk in the output graph.

We do so in a process based on three complementary graph models.
(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \citep{Kehr_2014}.
(\textit{2}) We then contract match components in this graph into single nodes of a \textit{sequence graph} \citep{Hein_1989}.
(\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
In doing so, we are able to derive edges and label the paths of a \textit{variation graph} \citep{Garrison_2018} which captures the pairwise alignments and all implied transitive relationships between them.

@@

As input, we have $Q = g_1 \ldots g_N$, which is a concatenation of the $N$ genome sequences from which we will build the graph.
Using a compressed suffix array (CSA) over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.

@@

>>>>

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

\section{Variation graph induction}
\label{sec:algorithm}

A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
A \textit{bidirected} sequence graph represents both strands of DNA.
On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Eizenga_2020}.

\begin{definition}
\label{def:vg}
Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2019_thesis}.
In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
A node strand $s$ corresponds to one node orientation.
Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.
\end{definition}

\begin{theorem}
A variation graph represents pairwise alignments between its embedded paths.
\end{theorem}

\begin{proof}
By definition \ref{def:vg}, two paths have identical subsequences where they walk (or ``step'') through the same series of oriented nodes (e.g. $s_1 s_2 s_3$).
An identical set of path steps is thus equivalent to a sequence match.
Pairwise alignments are by definition collections of matches between sequences.
The variation graph thus models a set of pairwise alignments between paths in $P$.
\end{proof}

\begin{theorem}
We can construct a variation graph from pairwise alignments.
\end{theorem}

This follows from \ref{def:vg}.
Our input $Q = S \vee \bar{S}$ is a set of $N$ DNA sequences $S = g_1 \ldots g_N$ and their reverse complements $\bar{S} = \bar{g_1} \ldots \bar{g_N}$.
%Assume that $i$ and $j$ are coordinates in $Q$.
A match $m = (i, j)$ asserts the evolutionary equivalence of two charaters in sequences in $Q$.
%$\forall_{(i, j) : j = |g| - i} g[i] = \bar{g}[j]$.
Pairwise alignments between sequences in $Q$ are a set of matches $A = \{ m_1 \ldots m_{|A|} \}$.
By standard definition, each sequence matches its own reverse complement: $g[i] = \bar{g}[j]$ for all $j = |g| - i$, and we assume these matches are included in $A$.
%A match $m$ can be understood as a component in a binary relation.
The transitive closure of a match, $m^+ = \{ i \ldots j \}$, is a set of characters in $Q$ that are transitively linked together by other matches.
By definition of $m$, each $m^+$ implies a single, identical character $c(m^+)$.

We build a graph $V$ inductively.
We take the first match in $A$, $m_1$, and execute a union-find operation to obtain $m_1^+$.
We add the character of the match $c(m_1^+)$ as a node $n_1$ in $V$, record the mapping from $m_1^+ \to n_1$.
To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $V$ while recording the match transitive closure to graph mapping.
We continue until all matches have been used.
Finally, we establish paths ($P$) by projecting $Q$ through the match-to-node mapping, and record edges ($E$) where nodes occur successively in paths.

\begin{proof}
After the first step of induction, the graph represents all pairwise matches in $m_1^+$.
Each subsequent step includes progressively more of $A$, until at completion, all pairwise relationships are accounted for in $V$.
The variation graph is constructed from pairwise alignments, and because its paths are equivalent to input sequences, it represents them losslessly.
\end{proof}

\begin{comment}

Our algorithm flows through three complementary graph models.
(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \citep{Kehr_2014}.
(\textit{2}) We then transitively close match relations in this graph to produce single nodes of a \textit{sequence graph} \citep{Hein_1989}.
(\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
In doing so, we are able to derive edges and label the paths of a \textit{variation graph} \citep{Garrison_2018} which captures the pairwise alignments and all implied transitive relationships between them.

@@

Using a compressed suffix array (CSA) over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.

@@

\end{comment}
